Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-30 tagless/tagless-final.tex
a few cosmetic changes

The main tool is to take functions seriously: we encode the
higher-order abstract syntax of object terms not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
Our family of evaluators contains an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), which lets us encode DSL terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.
\oleg{say something about type functions, GADTs and the way to get
  around them?}\jacques{not in the abstract?}
\end{abstract}

@@

    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
both
interpreter of a typed language in a typed language is an interesting
problem. The known solutions include the use of the Universal type, or
GADTs. The former is not quite satisfactory for performance reasons;
and also because of the presence of partial pattern matching in the
interpreter; the inexhaustiveness of the pattern matching should never
be triggered when executing a well-typed term. Alas, this fact cannot
be made obvious to the compiler. That also exlains unoptimality.

@@

and a sample term is
\begin{code}
  let test1 = A ((L (V VZ)),(B true))
\end{code}

Let us attempt to implment an the interpreter, the function |eval0|. It takes
an object language term such as |test1| and gives us its value.
The first argument to |eval| is the environment, initially empty,
which is the sequence of values associated with free variables in the
interpreted code.
\begin{code}
  let rec eval0 env = function 
  | V VZ      -> List.hd env
  | V (VS v)  -> eval0 (List.tl env) (V v)
  | B b       -> b 
  | L e       -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

Suppose our meta-language (the language in which we wrote test1
and that interpreter) were untyped. The above code would be acceptable.
The |L e| line exhibits interpretive overhead \cite{WalidICFP02}: 
|eval0 env' e| will be executed every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead (sections 1.1 and 1.2 of \cite{WalidICFP02} explain
this well).

If we use OCaml or some other typed language as the meta-language, 
the function |eval0| is ill-typed. The line |B b|
says that the function returns a boolean, whereas the next line says
the result is a function. Clearly, the result of |eval0|
can't be both of these types. 
A related problem is the type of the environment |env|: since variables
in the language may be either of boolean or function types, a regular
OCaml list cannot be used for holding the corresponding values.

@@

Unfortunately, the result is tagged: |UB true| rather than just
|true|. We also note that the eval function is partial in two
respects: first there is an inexhaustive pattern-matching in the 
line |A (e1,e2)|.
Second, the occurrences of |List.hd| and |List.tl|.
The failure of the latter corresponds to the evaluation of `open'
terms. One can see that if all our terms are closed, |List.hd| and
|List.tl| above will never be applied to the empty list. The partial
pattern match in the |A (e1,e2)| line raises an exception when we try to
evaluate something like |A (B true, B false)|, that is when the first
operand is not a function.

@@

\begin{code}
  let b (bv:bool) env  = bv
  let varZ (vc,_)      = vc
  let varS vp (_,envr) = vp envr
  let app e1 e2 env    = (e1 env) (e2 env)
  let lam e env        = fun x -> e (x,env)
\end{code}

@@

In the above code, the interpreter is wired in, directly built into the
functions |b|, |app|, etc.  In the remainder of this paper, 
we explain how to abstract the interpreter,
to make it possible to ``evaluate'' the same term in many different
ways: interpretation, compilation, compute the size (or depth), etc.

@@

We have also embedded our DSL in (Meta)OCaml, replacing type classes with
functors, via a module type.  For this section, the following simple
signature will be sufficient:

@@

This interpreter is patently tagless: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. We can run our
examples by instatiating the |EX| functor with |R|.
\noindent
\begin{code}
  module EXR = EX(R);;
\end{code}

\noindent and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define a\\
|newtype R a = R a| and |instance Symantics R|
with a straightforward implementation. We must stress that |R| is
a |newtype|. Although it may look like a tag, it has no run-time
representation (erased during compilation) and pattern-matching
against |R| cannot ever fail and so is operationally an identity and
compiled that way.

Note that the interpreter uses no Universal type nor pattern
matching, and it is obvious to the compiler that non-termination can
only result from interpreting |fix|.

\begin{prop}
If the term |e| in the meta-language has type
|t|, evaluation of that term in the interpreter R will either continue
indefinitely or termninate with a value of the same type |t|.
\end{prop}

@@

\subsection{The trouble with (opaque) functions}
\label{S:PE-problem}
We want to combine the interpreter |R| and the compiler |C| to produce
an (online) partial evaluator.  We defer the discussion of the 
choice of online/offline partial evaluation until after we have presented our
solution in~\ref{S:PE-solution}.
The partial evaluator (PE) is also an interpreter: it
evaluates a term either to an immediate (or, `static' or present-stage)
value or to a `dynamic', future-stage value, or `code' (of the sort
produced by the compiler C). The interpreter attempts to perform all
possible computations (i.e., computations whose operarands are
present-stage values) as the present-stage. Our first attempt is
straightforward: we introduce
\begin{code}
data P1 t = S1 (R t) (C t) | E1 (C t)
\end{code}
as a `direct sum' of the interpreter |R| and the compiler |C|. The
datatype |P1| has two \emph{phase} (not type!) tags |S1| 
(for static) and |E1| for dynamic. The static value has both the
present-stage and future-stage alternatives. The dynamic value does
not have the present-stage component. We define
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (S1 _ dyn) = dyn
abstr1 (E1 dyn)   = dyn
\end{code}
to extract the future-stage component. We then attempt to define the
interpreter |P1| compositionally -- and it works for the first-order
fragment of our DSL language.

@@

When we come to the higher-order fragment of our language, however, we
stumble. Let us consider |lam f|. Here, |f| is a function from a
|P1 a| value to a |P1 b| value. The result of |lam f| must be either a
static value or a dynamic value -- but which? Unlike |add| or |if_|, 
|lam| cannot really examine its operand. The arguments of |add| are 
of type |P1 int|, which we can examine by pattern-matching and so 
determine if they are static or dynamic. The argument of |lam| is a 
higher-order function, which is not introspectible. One may only 
apply |f| to a particular static or dynamic value, and receive another 
value in return. Alas, as in the case |lam (\x -> mul x dynval)|, 
applying the same function to different static values may give 
either static or dynamic results (in the example above, applying 
the function to (int 0) gives the static value in return). One
solution is to be conservative and always use the |C| compiler to
compile |f| to the future-stage expression, which we can tag with
E1. But that defeats the purpose of partial evaluation as we may miss
many opportunities for static computations and simplifications, such
as multiplication by zero in the example above. Another solution is to
delay the binding-type-analysis (BTA) until the function |f| is
applied to a particular argument (that is, until the result of |lam f|
appears as the argument of |app|). That means, we have to find a way
to incorporate |f| as it is into the P1 data structure. We would like
to write that when |t| is a particular type |a->b|, that data type
should be 
\begin{code}
data P1 (a->b) = S1 (R (P1 a -> P1 b)) (C (a->b)) 
               | E1 (C (a->b))
\end{code}
That is, we need a non-parameteric data type, something like a
typecase (ref? Gibbons?).

The problem we encountered here is quite akin to typed CPS
transform. When transforming abstractions, we come across
non-factorialness. The connection with CPS is not an accident, as we
shall see below (in Section XXX).

@@

Two common ways of providing typecase in Haskell is by using either
generalized algebraic data types (GADT) or type-class functions
(implemented via typeclasses with functional dependencies). These
methods are equivalent; here we show the GADT; please see |incope1.ml|
in the accompanying source code for the latter.

@@

\subsection{The `final' solution}
\label{S:PE-solution}
Let us re-examone the problem in Section~\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = S (repr_pe t) (C t) | E (C t)
\end{code}
where |repr_pe| is a type function, defined inductively as 
\begin{code}
repr_pe Int    = Int
repr_pe Bool   = Bool
repr_pe (a->b) = (repr_pr a) -> (repr_pe b)
\end{code}
Although we can use typeclasses to define that type function
in Haskell, that is not portable to MetaOCaml. We notice however that
the type case alternatives above are all associated with the existing
methods of |Symantics| and so the case analysis is already present. 
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml for
portability, we recall from Section~\ref{encoding} that |repr| type
had two arguments |type ('c,'dv) repr|. We add an extra argument,
|'sv|, which is the result of applying |repr_pe| function above to the
type |'dv|. The signature |Symantics| now reads
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int  : int  -> ('c,int,int) repr
  val bool : bool -> ('c,bool,bool) repr
  val add  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val mul  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val leq  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,bool,bool) repr
  val if_  : ('c,bool,bool) repr ->
             (unit -> ('c,'sa,'da) repr) ->
             (unit -> ('c,'sa,'da) repr) -> 
             ('c,'sa,'da) repr 
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
           'da->'db) repr
  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
             'da->'db) repr
    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
  val fix : 
    (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,
         'da->'db) repr 
    -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
end;;
\end{code}
The interpreters |R|, |L| and |C| above only relied on the last
argument of the |repr| type, which is unchanged. And so is their
implementation; the newly added argument |'sv| will be phantom.
For example, we would have write, in the case of the |C| interpreter,
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
When writing the partial evaluator however, we make use of that
type argument |'sv|. Here's the corresponding instance of the
Symantics:

@@

\subsection{Translucent types}
The crucial role of the higher-order type parameter r

@@

>>>>

The main tool is to take functions seriously: we encode the
higher-order abstract syntax of object terms not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of final algebras.
Our family of evaluators contains an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), which lets us encode DSL terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.
\end{abstract}

@@

    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
both
interpreter of a typed language in a typed language is an interesting
problem. The known solutions include the use of the Universal type, or
GADTs. The former is not quite satisfactory for performance reasons;
and also because of the presence of partial pattern matching in the
interpreter; the inexhaustiveness of the pattern matching should never
be triggered when executing a well-typed term. Alas, this fact cannot
be made obvious to the compiler. That also exlains unoptimality.
\oleg{Should say somewhere here that our code is surprisingly simple,
  and, in the hindsight, obvious. The problem of typed direct
  intrepreter of a typed language is difficult and thought to require
  GADT or dependent types. 
  In fact, several papers introducing GADTs (Xi03 and SPJ) as well as 
  (language in Walid03) and Concoqtion, written 2003-2007, used the 
  typed tagless intrepreter as the motivating example.
  Taha (in Jones optimality paper, I think) said that typed
  interpreter of the typed language was not solved then.
  Or, should we refer to Sec `Other solutions'}

@@

with a sample term (application of the identity function to a boolean) as
\begin{code}
  let test1 = A ((L (V VZ)),(B true))
\end{code}

Let us attempt to implement an the interpreter, the function |eval0|. It takes
an object language term such as |test1| and gives us its value.
The first argument to |eval| is the environment, initially empty,
which is the sequence of values associated with free variables in the
interpreted code.
\begin{code}
  let rec eval0 env = function 
  | V VZ      -> List.hd env
  | V (VS v)  -> eval0 (List.tl env) (V v)
  | B b       -> b 
  | L e       -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

Suppose our meta-language (the language in which we wrote |test1|
and that interpreter) were untyped. The above code would be acceptable.
The |L e| line exhibits interpretive overhead \cite{WalidICFP02}: 
|eval0 env' e| will be executed every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead (sections 1.1 and 1.2 of \cite{WalidICFP02} explain
this well).

If we use OCaml or some other typed language as the meta-language, 
the function |eval0| is ill-typed. The line |B b|
says that |eval0| returns a boolean, whereas the next line says
the result is a function. All branches of a pattern-match form must
yield values of the same type, however. 
A related problem is the type of the environment |env|: since variables
in the language may be either of boolean or function types, a regular
OCaml list cannot be used for holding the corresponding values.

@@

Unfortunately, the result is tagged: |UB true| rather than just
|true|. We also note that the eval function is partial in two
respects: first there is an inexhaustive pattern-matching 
|match ... with UA f -> ...| in the 
line |A (e1,e2)|.
Second, the occurrences of |List.hd| and |List.tl|.
The failure of the latter corresponds to the evaluation of `open'
terms. One can see that if all our terms are closed, |List.hd| and
|List.tl| above will never be applied to the empty list. The partial
pattern match in the |A (e1,e2)| line raises an exception when we try to
evaluate something like |A (B true, B false)|, that is when the first
operand is not a function.

@@

\begin{code}
  let b (bv:bool) env  = bv
  let varZ env         = fst env
  let varS vp env      = vp (snd env)
  let app e1 e2 env    = (e1 env) (e2 env)
  let lam e env        = fun x -> e (x,env)
\end{code}

@@

In the above code, the interpreter is wired in, directly built into the
functions |b|, |app|, etc.  In the remainder of this paper, 
we explain how to abstract the interpreter,
to make it possible to ``evaluate'' the same term in many different
ways: interpretation, compilation, partial evaluation, size
computation, etc.

@@

We have also embedded our DSL in (Meta)OCaml, replacing type classes
and instances with signatures and structures. For this and the 
following sections, the following simple signature suffices:

@@

This interpreter is patently tagless, using neither Universal type nor
any pattern-matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. We can run our
examples by instatiating the |EX| functor with |R|.
\noindent
\begin{code}
  module EXR = EX(R);;
\end{code}

\noindent and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define a\\
|newtype R a = R{unR::a}| and |instance Symantics R|
with a straightforward implementation. We must stress that |R| is
a |newtype|. Although it may look like a tag, it has no run-time
representation (erased during compilation) and pattern-matching
against |R| cannot ever fail and so is operationally an identity and
compiled that way. In both cases, the absense of pattern-match
failures is obvious to the compiler, because there are no
pattern-matching. The failure to yield a value
may only result from interpreting |fix|.

\begin{prop}
If the term |e| encoded in the meta-language has type
|t|, evaluation of that term in the interpreter R will either continue
indefinitely or termninate with a value of the same type |t|.
\end{prop}

@@

\subsection{The trouble with (opaque) functions}
\label{S:PE-problem}
We want to combine the interpreter |R| and the compiler |C| to produce
an (online) partial evaluator.  We defer the discussion of the 
choice of online/offline partial evaluation until after we have presented our
solution in~\ref{S:PE-solution}.
The partial evaluator (PE) is too an interpreter: it
evaluates a term either to an immediate (or, `static' or present-stage)
value or to a `dynamic', future-stage value, or `code' (of the sort
produced by the compiler C). The interpreter attempts to perform all
possible computations (i.e., computations whose operarands are
present-stage values) as the present-stage. Our first attempt, in
Haskell, is straightforward: we introduce
\begin{code}
data P1 t = S1 (R t) (C t) | E1 (C t)
\end{code}
as a `direct sum' of the interpreter |R| and the compiler |C|. The
datatype |P1| has two \emph{phase} (not type!) tags |S1| 
(for static) and |E1| for dynamic. The static value has both the
present-stage and future-stage alternatives. The dynamic value does
not have the present-stage component. We define
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (S1 _ dyn) = dyn
abstr1 (E1 dyn)   = dyn
\end{code}
to extract the future-stage component. We then attempt to define the
interpreter |P1| compositionally -- and it works for the first-order
fragment of our DSL language.

@@

When we come to the higher-order fragment of our language, however, we
stumble. Let us consider |lam f|. Here, |f| is a function from a
|P1 a| value to a |P1 b| value. The result of |lam f| must be either a
static value or a dynamic value -- but which? Unlike |add| or |if_|, 
|lam| cannot really examine its operand. The arguments of |add| are 
of type |P1 int|, which we can examine by pattern-matching and so 
determine if they are static or dynamic. The argument of |lam| is a 
function, which is not introspectible. One may only 
apply |f| to a particular static or dynamic value, and receive another 
value in return. Alas, as in the case |lam (\x -> mul x dynval)|, 
applying the same function to different static values may give 
either static or dynamic results (in the example above, applying 
the function to (int 0) gives the static value in return). One
solution is to be conservative and always use the |C| compiler to
compile |f| to the future-stage expression, which we can tag with
E1. But that defeats the purpose of partial evaluation as we may miss
many opportunities for static computations and simplifications, such
as multiplication by zero in the example above. A better solution is to
delay the binding-type-analysis (BTA) until the function |f| is
applied to a particular argument (that is, until the result of |lam f|
appears as the argument of |app|). That means, we have to find a way
to incorporate |f| as it is into the P1 data structure. We would like
to write that when |t| is a particular type |a->b|, that data type
should be 
\begin{code}
data P1 (a->b) = S1 (R (P1 a -> P1 b)) (C (a->b)) 
               | E1 (C (a->b))
\end{code}
That is, we need a non-parameteric data type, something like a
typecase (ref? Gibbons?).

The problem we encountered here is quite akin to typed CPS
transform. When transforming abstractions, we come across
non-functorialness. The connection with CPS is not an accident, as we
shall see below (in Section XXX).

@@

Two common ways of providing typecase in Haskell is by using either
generalized algebraic data types (GADT) or type-class functions
(implemented via typeclasses with functional dependencies). These
methods are equivalent; here we show the GADT; please see |incope1.hs|
in the accompanying source code for the latter.

@@

\subsection{The `final' solution}
\label{S:PE-solution}
Let us re-examone the problem in Section~\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = S (repr_pe t) (C t) | E (C t)
\end{code}
where |repr_pe| is a type function, defined inductively
% inductively because P below depends on repr_pe
as 
\begin{code}
repr_pe Int    = Int
repr_pe Bool   = Bool
repr_pe (a->b) = P a -> P b
\end{code}
Although we can use typeclasses to define that type function
in Haskell, that is not portable to MetaOCaml. We notice however that
the type case alternatives above are all associated with the existing
methods of |Symantics| and so the case analysis is already present. 
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml for
portability, we recall from Section~\ref{encoding} that |repr| type
had two arguments |type ('c,'dv) repr|. We add an extra argument,
|'sv|, which is the result of applying |repr_pe| function above to the
type |'dv|. The signature |Symantics| now reads
\begin{code}
module type Symantics = sig
  type ('c,'sv,'dv) repr
  val int  : int  -> ('c,int,int) repr
  val bool : bool -> ('c,bool,bool) repr
  val add  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val mul  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,int,int) repr
  val leq  : ('c,int,int) repr -> 
      ('c,int,int) repr -> ('c,bool,bool) repr
  val if_  : ('c,bool,bool) repr ->
             (unit -> ('c,'sa,'da) repr) ->
             (unit -> ('c,'sa,'da) repr) -> 
             ('c,'sa,'da) repr 
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
           'da->'db) repr
  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
             'da->'db) repr
    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
  val fix : 
    (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,
         'da->'db) repr 
    -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
end;;
\end{code}
The interpreters |R|, |L| and |C| above only relied on the last
argument of the |repr| type, which is unchanged. And so is their
implementation; the newly added argument |'sv| will be phantom.
For example, we would now write the interpreter |C| as
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
When writing the partial evaluator however, we make use of the
type argument |'sv|. Here's the corresponding instance of the
Symantics:

@@

\subsection{Translucent types}

The crucial role of the higher-order type parameter r

@@

The attempt to encode self-interpreter brings the |*->*| polymorphism
the the forefront, as we shall now see.
